\documentclass[conc-doc]{subfiles}

\newpage
\thispagestyle{plain} % empty
\mbox{}

\newpage
~\vfill
\thispagestyle{empty}

'Goddamn,' he said. He knew he sounded groggy with the cold and with the drug. 'An atheist my entire fucking life, and it turns out the credulous assholes were right all along!' He wheezed, coughed. 'Or do you surprise them too by not turning up?'\\
'You flatter me, Mr Zakalwe,' the woman said, in a superbly deep and sexy voice. 'I am not Death, or some imagined Goddess. I am as real as you...' She stroked his torn, bleeding palm with one long, strong thumb. 'If a little warmer.'\\
'Oh, I'm sure you're real,' he said. 'I can feel you're rea...'\\
His voice faded; he looked behind the woman. There was a huge shape appearing inside the whirling snow. Grey-white like the snow, but a single shade darker, it floated up behind the woman, quiet and huge and steady. The storm seemed to die, just around them.\\
'That's called a twelve person module, Cheradenine,' the woman said. 'It's come to take you away, if you want to be taken away; to the mainland, if you like. Or further afield, away with us if you'd prefer that.'\\
He was tired of blinking and shaking his head. Whatever insane part of his mind wanted to play this bizarre game out would just have to be humoured for as long as it took. What it had to do with the Staberinde and the Chair, he couldn't tell yet, but if that was what it was all about - and what else could it be about? - then there was still no point, in this weakened, dying state, trying to fight it. Let it happen. He had no real choice. 'With you?' he said, trying not to laugh.\\
'With us. We'd like to offer you a job.' She smiled. 'But let's talk somewhere a little warmer, shall we?'\\

\begin{verse}
\attribute{
	\textsc{Iain M. Banks (2008)}\\
	\textit{Use of Weapons}\\
}
\end{verse}

\pagestyle{fancy} % Enable headers and footers again

\newpage
\thispagestyle{plain} % empty
\mbox{}